Live Functional Programming with Typed Holes

### 2018-04-11 calculus.tex
added line breaks at sentence boundaries in sec 3 and addressed ravis comments in blue, along with a few other small tweaks. ready for Matt to give it a pass.

We will now make the intuitions developed in the previous section formally precise by specifying the \HazelnutLive core calculus and  its accompanying metatheory. We have mechanized these formal developments using the Agda proof assistant \cite{norell:thesis,norell2009dependently} (see Sec.~\ref{sec:agda-mechanization} for some additional details, and the supplemental material for the full mechanization). % The proofs of the metatheorems are established by this mechanization; the proofs are only very briefly outlined in this section.

@@

We give an overview of the static semantics of the \HazelnutLive
external language.

\Figref{fig:bidirectional-typing} defines \emph{bidirectional} typing
%
as two mutually defined judgements: The type synthesis
judgement~$\hsyn{\hGamma}{\hexp}{\htau}$ synthesizes a type~$\htau$
for external expression~$\hexp$ under typing context $\hGamma$, which tracks typing
assumptions of the form $x : \htau$ in the usual
manner \cite{pfpl,tapl}.
%
The type analysis judgement~$\hana{\hGamma}{\hexp}{\htau}$ checks
expression~$\hexp$ against a given type~$\htau$.
%
Algorithmically, analysis accepts a type as input, and synthesis gives
a type as output.
%
We use synthesis for each of the programmer's ``top level'' external
exressions~$e$, giving the corresponding type as feedback.

@@

The benefit of specifying the \HazelnutLive external language
bidirectionally is that the programmer does not need to annotate each
hole with a type
%
\cite{Pierce:2000ve,bidi-tutorial,DBLP:conf/icfp/DunfieldK13,Chlipala:2005da}.
%
These typing rules closely follow the \Hazelnut type
system \cite{popl-paper}; we discuss minor differences below.

We supplement each expression hole from \Hazelnut with a
hole \emph{name}, which are new to \HazelnutLive; an empty hole is
written $\hehole{u}$, where $u$ is a unique name for the hole.
%
Rule {SEHole}\todo{rule name macros}{} specifies that an empty hole synthesizes hole type, written $\tehole$.
%
If an empty hole appears where an expression of some other type is
expected, e.g. under an explicit ascription (governed by Rule {SAsc})
or in the argument position of a function application (governed by
Rule {SAp}, discussed below), we apply the \emph{subsumption rule},
Rule {ASubsume}, which specifies that if an expression $e$ synthesizes
type $\htau$, then it can be checked against any \emph{consistent}
type, $\htau'$.

@@

For the familiar forms of the lambda calculus, the rules again follow prior work. For simplicity, the core calculus includes only a single base type, $b$, with a single constant, $c$, governed by Rule {SConst} (i.e. $b$ is the unit type). \Hazelnut instead defined a number type with a single operation, which we include in Appendix~\ref{sec:extensions} alongside various other standard extensions to the core calculus\todo{do this, say more?}. 
%
\rkc{perhaps too much info / unnecessary to say Hazelnut had numbers instead of unit?} 
%
Rule {SVar} specifies that variables synthesize the corresponding type from $\hGamma$.

@@

The rule governing function application, Rule {SAp}, similarly treats expressions of hole type in function position as if they were of type $\tarr{\tehole}{\tehole}$ using the same matched arrow type judgement.

\subsection{Expansion}
\label{sec:expansion}

@@

Expansion is specified, much like the type system for the external language, in the bidirectional style by two mutually defined judgements. The synthetic expansion judgement, $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$, synthesizes a type, $\htau$, from $\hexp$, and produces an expansion, $d$, and a hole context, $\hDelta$. We say more about hole contexts below. The analytic expansion judgement, $\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$, checks $\hexp$ against $\htau$ and produces an expansion, $d$, of type $\htau'$, and a hole context, $\hDelta$. The governing theorem below establishes that the type $\htau'$ is necessarily consistent with $\htau$.
%
\rkc{minor, but the syntax of type assignment isn't mentioned before use. it's not
entirely standard, because of the hole context, so perhaps worth mentioning.}
%
\begin{thm}[Typed Expansion]\label{thm:typed-expansion} ~
  \begin{enumerate}[nolistsep]
    \item
      If $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$
      then $\hasType{\Delta}{\hGamma}{\dexp}{\htau}$.
    \item
      If $\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$
      then $\tconsistent{\htau}{\htau'}$ and $\hasType{\Delta}{\hGamma}{\dexp}{\htau'}$.
  \end{enumerate}
\end{thm}
\noindent
The reason analytic expansion produces an expansion of consistent type is because the subsumption rule, as previously discussed, allows us to check an external expression against any type consistent with the type the expression actually synthesizes, whereas every internal expression can be assigned at most one type, i.e. the following standard unicity property holds of the type assignment system: 
\begin{thm}[Type Assignment Unicity]
  If $\hasType{\Delta}{\hGamma}{\dexp}{\htau}$
  and $\hasType{\Delta}{\hGamma}{\dexp}{\htau'}$
  then $\htau=\htau'$.
\end{thm}
\noindent
Consequently, analytic expansion reports the type actually assigned to the expansion it produces. For example, we can derive that $\expandAna{\hGamma}{c}{\tehole}{c}{b}{\emptyset}$.% where $\emptyset$ is the empty hole context.

@@

The rules governing expansion of constants, variables and lambda expressions --- Rules {ESConst}, {ESVar}, {ESLam} and {EALam} --- and the corresponding type assignment rules --- Rules {TAConst}, {TAVar} and {TALam} --- mirror the typing rules from Fig.~\ref{fig:bidirectional-typing} (so the corresponding cases of Theorem~\ref{thm:typed-expansion}, Theorem~\ref{thm:expandability} and Theorem~\ref{thm:expansion-generality} are straightforward). Note that in the internal language, all lambdas are half-annotated, again to support type assignment---Rule {EALam} inserts the annotation onto unannotated external lambdas based on the given type. The rules governing expansion of holes, function application and ascription are more interesting, so let us consider them in turn.
%
\rkc{this list (holes, func app, and ascription) has three items, but there are
only two subsections. see comment at beginning of Sec 3.2.2.}

\subsubsection{Hole Expansion}\label{sec:hole-expansion} Rules {ESEHole}, {ESNEHole}, {EAEHole} and {EANEHole} govern the expansion of empty and non-empty expression holes to \emph{hole closures}, $\dehole{u}{\sigma}{}$ and $\dhole{\dexp}{u}{\sigma}{}$, respectively. 
%
\rkc{perhaps add ``empty and non-empty'' before hole closures.}

The hole name, $u$, on a hole closure identifies the external hole that the hole closure corresponds to. Note that while we assume each hole name to be unique in the external language, there can be multiple hole closures with the same name during evaluation due to substitution (though in the initial expansion, the uniqueness condition happens to hold because evaluation has yet to occur). For example, the result from Fig.~\ref{fig:grades-example} showed four closures for the hole named 1\todo{update with actual number (four?) and hole name later}. There, we numbered each hole closure for a given hole sequentially, \li{1:1}, \li{1:2} and so on, but this is strictly for the sake of presentation so we omit hole closure numbers from the core calculus.

The hole expansion rules are the only rules that introduce hypotheses, of the form $\Dbinding{u}{\hGamma}{\htau}$, into the hole context, $\Delta$. The purpose of the hole context is to record a type, $\tau$, and a typing context, $\Gamma$,   for each hole name, $u$.\footnote{We use a hole context, rather than recording the typing context and type directly on each hole closure, to ensure that all closures for a hole name have the same typing context and type.} This notation for hole contexts is taken from contextual modal type theory (CMTT) \cite{Nanevski2008}, identifying hole names with metavariables and hole contexts with modal contexts (we say more about the connection with CMTT below). In all four hole expansion rules, the typing context recorded in the hole context is simply the current typing context when the hole is expanded. In the synthetic hole expansion rules, {ESEHole} and {ESNEHole}, the generated hole context assigns the hole type, $\tehole$, to $u$, as in the typing rules. However, the first two premises of the expansion subsumption rule, Rule EASubsume, disallow the use of subsumption for holes in analytic position. Instead, we have separate analytic rules, {EAEHole} and {EANEHole}, which record the type that the hole is being checked against into the hole context. This is again so that we can use type assignment for the internal language --- the type assignment rules TAEHole and TANEHole in Fig.~\ref{fig:hasType} assign a hole closure for $u$ the corresponding type from the hole context.

@@

Empty hole closures, $\dehole{u}{\sigma}{}$,  correspond to the metavariable closures (a.k.a. deferred substitutions) from CMTT, $\cmttclo{u}{\sigma}$.\todo{change notation from CMTT}{} We will see how closure environments evolve during evaluation in Sec.~\ref{sec:evaluation}. Non-empty hole closures, $\dhole{d}{u}{\sigma}{}$, do not directly correspond to a notion from CMTT\todo{mention reason in Sec 4}.

\subsubsection{Cast Insertion}\label{sec:cast-insertion} 
%
\rkc{consider renaming subsection to ``Function Application and Cast
Intersion''. and adding a transition sentence like ``The remaining development
of expansion deals with the delicate interplay between function application and
cast insertion.''}
%
Consider the following example: $\hap{(\halam{x}{\tehole}{\hap{x}{c}})}{c}$. The type synthesized for this example viewed as an external expression is $\tehole$, because the hole type annotation on $x$ allows us to apply it as a function of type $\tarr{\tehole}{\tehole}$, as previously discussed, and $c$ can be checked against type $\tehole$ by subsumption. However, viewed as an internal expression, this example is not well-typed---we do not have subsumption in the type assignment system defined in Fig.~\ref{fig:hasType}. Indeed, it would violate type safety if we could assign a type to this example in the internal language, because beta reduction of this example viewed as an internal expression would result in $c(c)$, which is clearly not well-typed. The difficulty is that leaving the argument type unknown leaves how the argument is being used (in this case, as a function) also unknown.\footnote{In a system where type reconstruction is first used to try to fill in type holes, we could express a similar example by using $x$ at two or more different types, thereby causing type reconstruction to fail.
%On the other hand, if it is acceptable to arbitrarily choose one of the possible types, and type reconstruction is complete, then type holes will never appear in the internal language and the cast insertion machinery described in this section can be omitted entirely, leaving only the hole closure machinery described previously.
}\todo{talk about elsewhere? maybe do type-hole-free version of calculus in appendix if there is time}{} Recall that we can interpret hole types as unknown types from gradual type theory. This leads us to the solution to this problem: cast insertion. 
%
\rkc{Instead of the last two sentences, maybe something like: ``By analogy
between hole types in our system and unknown types from gradual type theory, we
address the problem with the same mechanism used in gradual type theory: cast
insertion.''}

The cast form in Hazelnut Live is $\dcasttwo{\dexp}{\htau_1}{\htau_2}$, which, as specified by Rule TACast in Fig.~\ref{fig:hasType}, serves to ``box'' an expression of type $\htau_1$ for treatment as an expression of some consistent type $\htau_2$.\footnote{In the earliest work on gradual type theory, the cast form only gave the target type, $\htau_2$ \cite{Siek06a}, but it simplifies matters to include the assigned type, $\htau_1$, in the syntax \cite{DBLP:conf/snapl/SiekVCB15}.}

Casts are inserted during the expansion of function applications and ascriptions. The latter is more straightforward: Rule~{ESAsc} in Fig.~\ref{fig:expandSyn} inserts a cast from the assigned type to the ascribed  type. Theorem~\ref{thm:typed-expansion} inductively ensures that the two types are consistent.  Note that we included ascription only for the sake of exposition: it can be defined using application together with the half-annotated identity function: $e : \tau = \hap{(\halam{x}{\htau}{x})}{e}$.
%
\rkc{move this Note sentence to the same place where half-annotated lambdas were
noted. This note breaks the flow here.}

Cast insertion by the application expansion rule, Rule~{ESAp}, has much the same intuition, but requires a bit more care. To understand the rule, consider the expansion for the example above:
\[\hap{\dcasttwo{(\halam{x}{\tehole}{\underline{\hap{\dcasttwo{x}{\tehole}{\tarr{\tehole}{\tehole}}}{\dcasttwo{c}{b}{\tehole}}}})}{\tarr{\tehole}{\tehole}}{\tarr{\tehole}{\tehole}}}{\dcasttwo{c}{b}{\tehole}}
\]
Let us focus on the expansion of the function body, underlined\todo{shaded?}, first. A cast is inserted on both the function expression, $x$, and the the argument, $c$.

The cast on $x$ allows us to treat the variable $x$, which is of type $\tehole$, as being of the matched arrow type $\tarr{\tehole}{\tehole}$, as described in Sec.~\ref{sec:external-statics}. The first three premises of Rule~{ESAp} accomplish this by first synthesizing a type for the function expression, here $\tehole$, then determining the matched arrow type, $\tarr{\tehole}{\tehole}$, and then performing analytic expansion on the function expression with this matched arrow type. The resulting expansion will have some type $\tau_1'$ consistent with the matched arrow type. In this case, because $x$ is of variable form, analytic expansion goes through subsumption so $\tau_1'$ is simply $\tehole$. The conclusion of the rule inserts the corresponding cast. Note that we go through type synthesis and analytic expansion so that the hole context records the matched arrow type for holes in function position, rather than the type $\tehole$ for all such holes as would be the case in a variant of this rule using synthetic expansion for the function expression\todo{put this in the appendix?}{}.

The conclusion of Rule~{ESAp} also inserts a cast on the argument's expansion, from the type it is assigned by the final premise of the rule, here $b$ as described at the start of Sec.~\ref{sec:expansion}, to the argument type, $\tehole$, of the matched arrow type of the function expression, here $\tehole$.

@@

The outer application in the example above goes through the same rule. In this case, the cast on the function is the identity cast for $\tarr{\tehole}{\tehole}$. We do not attempt to avoid the insertion of identity casts in the core calculus for simplicity (these will simply never fail during evaluation), but it is safe to do so (and some formal accounts of gradual typing do so by defining three application rules, including the original account of \cite{Siek06a}).

@@

The problem is first that \citet{Nanevski2008} defined only the logical reductions for CMTT, viewing it as a proof system for intuitionistic contextual modal logic via the propositions-as-types (Curry-Howard) principle. The paper therefore proved only a subject reduction property (which is closely related to type preservation). This is not a full dynamic semantics, and in particular, there is no notion of \emph{progress}, i.e. that well-typed terms cannot get ``stuck'' in an undefined state \cite{wright94:_type_soundness}. In any case, a conventional dynamic semantics for CMTT would not be immediately relevant to our goal of evaluating incomplete programs because, by our interpretation of hole closures, we would need a dynamic semantics for terms with free metavariables. \citet{Nanevski2008} sketched an interpretation of CMTT into the simply-typed lambda calculus with sums under permutation conversion\footnote{Permutation conversions are necessary to encode the commutative reductions of CMTT, which in turn are necessary to prove a strong normalization property. These issues are not relevant in \HazelnutLive because, as in the gradually typed lambda calculus, type holes admit non-termination: we can express the Y combinator as $(\halam{x}{\tehole}{x(x)}) (\halam{x}{\tehole}{x(x)})$.}, which has been studied by \citet{DBLP:journals/iandc/Groote02}, but under this interpretation an analagous problem arises---metavariables become variables of a function type, so again we cannot rely on any standard notion of progress on closed terms.\todo{citations}{}% We also cannot rely on, for example, weak head normalization because \HazelnutLive admits non-termination (due to casts).\todo{citation}

Furthermore, we need to integrate casts into the dynamic semantics. Fortunately, the dynamic semantics for the cast calculus from the gradually typed lambda calculus provides most, but not all, of the necessary machinery. The first problem is again with progress: in the cast calculus, the only irreducible terms of hole type are casts, which are accounted for by the progress theorem, but in $\HazelnutLive$, holes induce additional irreducible terms. The second missing piece is that in prior work on casts, evaluation aborts when cast failure occurs. Our goal, as discussed in Sec.~\ref{sec:failed-cast-example}, is for cast failure to instead insert a membrane around the dynamic type error, much like a non-empty hole serves as a membrane around a static type error, allowing in both cases for evaluation to safely and meaningfully continue past the error when desired.

The dynamic semantics for \HazelnutLive specified in Figures~\ref{fig:isGround}-\ref{fig:contextual-dynamics} addresses the difficulties just outlined, resulting in a system capable of running incomplete programs without aborting when a hole is encountered. This semantics is equipped with a meaningful notion of type safety (involving both preservation and progress). We also establish that the standard notion of type safety falls out when running complete terms. % We also establish that the standard type safety theorem holds when running complete terms.

The dynamic semantics is specified as a ``small-step'' transition system. The cast-related machinery is based closely on the cast calculus from the ``refined'' account of the gradually typed lambda calculus by \citet{DBLP:conf/snapl/SiekVCB15}, which is known to be theoretically well-behaved. In particular, Fig.~\ref{fig:isGround} defines the judgement $\isGround{\htau}$, which distinguishes the base type, $b$, and the least specific arrow type, $\tarr{\tehole}{\tehole}$, as \emph{ground types}, which play a role in simplifying the treatment of function casts. Fig.~\ref{fig:isFinal} defines the judgement $\isFinal{d}$, which distinguishes the final forms of the transition system. There are two classes of final forms: (possibly-)boxed values and indeterminate forms.\footnote{In most accounts of the cast calculus, values and ground types are distinguished with separate grammars together with an implicit identification convention. Our judgemental formulation is more faithful to the mechanization and cleaner for our purposes, because we are distinguishing several classes of final forms.}

@@

The judgement $\isBoxedValue{d}$ distinguishes (possibly-)boxed values, which correspond to the values from the cast calculus and include the classic values from the lambda calculus, distinguished by $\isValue{d}$, as well as two cast forms: casts between function types (except identity casts) and casts from a ground type to the hole type. In both cases, the cast must be on a boxed value. % (in the prior work, ground types and values were distinguished syntactically, but it is cleaner and more faithful to the mechanization to make these distinctions judgementally).

The judgement $\isIndet{d}$ distinguishes \emph{indeterminate} forms, so named because they arise from the presence of expression holes and failed casts in a program, and so, conceptually, their ultimate value awaits programmer action (see Sec.~\ref{sec:commutativity}). The first two rules specify that hole closures are indeterminate (in the case of non-empty hole closures, when the inner expression is final). The rules defining the remaining indeterminate forms are explained simultaneously with the corresponding transition rules below.\todo{relationship to weak head normal forms}

>>>>

We will now make the intuitions developed in the previous section formally precise by specifying the \HazelnutLive core calculus and  its accompanying metatheory. 
We have mechanized these formal developments using the Agda proof assistant \cite{norell:thesis,norell2009dependently} 
(see Sec.~\ref{sec:agda-mechanization} for some additional details, and the supplemental material for the full mechanization).

@@

We start with the type system of the \HazelnutLive
external language, which closely follows the \Hazelnut type
system \cite{popl-paper}; we discuss the minor differences as they come up below.

\Figref{fig:bidirectional-typing} defines the type system in the \emph{bidirectional} style 
%
with two mutually defined judgements \cite{Pierce:2000ve,bidi-tutorial,DBLP:conf/icfp/DunfieldK13,Chlipala:2005da}. The type synthesis
judgement~$\hsyn{\hGamma}{\hexp}{\htau}$ synthesizes a type~$\htau$
for external expression~$\hexp$ under typing context $\hGamma$, which tracks typing
assumptions of the form $x : \htau$ in the usual
manner \cite{pfpl,tapl}.
%
The type analysis judgement~$\hana{\hGamma}{\hexp}{\htau}$ checks
expression~$\hexp$ against a given type~$\htau$.
%
Algorithmically, analysis accepts a type as input, and synthesis gives
a type as output.
%
We start with synthesis for the programmer's ``top level'' external
exression.

@@

The primary benefit of specifying the \HazelnutLive external language 
bidirectionally is that the programmer does not need to annotate each hole with a type. 
%
An empty hole is
written $\hehole{u}$, where $u$ is a name for the hole, which we tacitly assume is unique. 
Holes did not have names in \Hazelnut. 
%
Rule {SEHole}\todo{rule name macros}{} specifies that an empty hole synthesizes hole type, written $\tehole$.
%
If an empty hole appears where an expression of some other type is
expected, e.g. under an explicit ascription (governed by Rule {SAsc})
or in the argument position of a function application (governed by
Rule {SAp}, discussed below), we apply the \emph{subsumption rule},
Rule {ASubsume}, which specifies that if an expression $e$ synthesizes
type $\htau$, then it can be checked against any \emph{consistent}
type, $\htau'$.

@@

For the familiar forms of the lambda calculus, the rules again follow prior work. For simplicity, the core calculus includes only a single base type, $b$, with a single constant, $c$, governed by Rule {SConst} (i.e. $b$ is the unit type). \Hazelnut instead defined a number type with a single operation, which we include in Appendix~\ref{sec:extensions} alongside various other standard extensions to the core calculus\todo{do this, say more?}. 
%
Rule {SVar} specifies that variables synthesize the corresponding type from $\hGamma$.

@@

The rule governing function application, Rule {SAp}, similarly treats an expression of hole type in function position as if it were of type $\tarr{\tehole}{\tehole}$ using the same matched arrow type judgement.

\vspace{-4px}
\subsection{Expansion}
\label{sec:expansion}
\vspace{-1px}

@@

Expansion is specified, much like the type system for the external language, in the bidirectional style by two mutually defined judgements. 
The synthetic expansion judgement, $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$, synthesizes a type, $\htau$, from $\hexp$, and produces an expansion, $d$, and a hole context, $\hDelta$. 
We say more about hole contexts, which are used in the type assignment judgement, $\hasType{\Delta}{\hGamma}{d}{\htau}$, below. 
The analytic expansion judgement, $\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$, checks $\hexp$ against $\htau$ and produces an expansion, $d$, of type $\htau'$, and a hole context, $\hDelta$. 
The governing theorem below establishes that expansions are well-typed and in the analytic case that the type $\htau'$ is necessarily consistent with $\htau$.
%
% \rkc{minor, but the syntax of type assignment isn't mentioned before use. it's not
% entirely standard, because of the hole context, so perhaps worth mentioning.}
%
\begin{thm}[Typed Expansion]\label{thm:typed-expansion} ~
  \begin{enumerate}[nolistsep]
    \item
      If $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$
      then $\hasType{\Delta}{\hGamma}{\dexp}{\htau}$.
    \item
      If $\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$
      then $\tconsistent{\htau}{\htau'}$ and $\hasType{\Delta}{\hGamma}{\dexp}{\htau'}$.
  \end{enumerate}
\end{thm}
\noindent
The reason analytic expansion produces an expansion of consistent type is because the subsumption rule, as previously discussed, allows us to check an external expression against any type consistent with the type the expression actually synthesizes, whereas every internal expression can be assigned at most one type, i.e. the following standard unicity property holds of the type assignment system.
\begin{thm}[Type Assignment Unicity]
  If $\hasType{\Delta}{\hGamma}{\dexp}{\htau}$
  and $\hasType{\Delta}{\hGamma}{\dexp}{\htau'}$
  then $\htau=\htau'$.
\end{thm}
\noindent
Consequently, analytic expansion reports the type actually assigned to the expansion it produces. For example, we can derive that $\expandAna{\hGamma}{c}{\tehole}{c}{b}{\emptyset}$.% where $\emptyset$ is the empty hole context.

@@

The rules governing expansion of constants, variables and lambda expressions --- Rules {ESConst}, {ESVar}, {ESLam} and {EALam} --- and the corresponding type assignment rules --- Rules {TAConst}, {TAVar} and {TALam} --- mirror the typing rules from Fig.~\ref{fig:bidirectional-typing} 
(so the corresponding cases of Theorem~\ref{thm:typed-expansion}, Theorem~\ref{thm:expandability} and Theorem~\ref{thm:expansion-generality} are straightforward). 
Note that in the internal language, all lambdas are half-annotated, again to support type assignment---Rule {EALam} inserts the annotation when expanding an unannotated external lambda based on the given type. 
The rules governing holes and the rules governing function applications and ascriptions, which both perform \emph{cast insertion}, are more interesting, so let us consider both cases in turn.

\subsubsection{Hole Expansion}\label{sec:hole-expansion} 
Rules {ESEHole}, {ESNEHole}, {EAEHole} and {EANEHole} govern the expansion of empty and non-empty expression holes to empty and non-empty \emph{hole closures}, $\dehole{u}{\sigma}{}$ and $\dhole{\dexp}{u}{\sigma}{}$. 
%

The hole name, $u$, on a hole closure identifies the external hole that the hole closure corresponds to. 
Note that while we assume each hole name to be unique in the external language, there can be multiple hole closures with the same name during evaluation due to substitution. 
For example, the result from Fig.~\ref{fig:grades-example} showed four closures for the hole named 1\todo{update with actual number (four?) and hole name later}. 
There, we numbered each hole closure for a given hole sequentially, \li{1:1}, \li{1:2} and so on, but this is strictly for the sake of presentation so we omit hole closure numbers from the core calculus.

The hole expansion rules are the only rules that introduce hypotheses, of the form $\Dbinding{u}{\hGamma}{\htau}$, into the hole context, $\Delta$. 
The purpose of the hole context is to record a type, $\tau$, and a typing context, $\Gamma$, for each hole name, $u$.\footnote{
We use a hole context, rather than recording the typing context and type directly on each hole closure, to ensure that all closures for a hole name have the same typing context and type.} 
This notation for hole contexts is taken from contextual modal type theory (CMTT) \cite{Nanevski2008}, identifying hole names with metavariables and hole contexts with modal contexts (we say more about the connection with CMTT below). 
In all four hole expansion rules, the typing context recorded in the hole context is simply the current typing context when the hole is expanded. 
In the synthetic hole expansion rules, {ESEHole} and {ESNEHole}, the generated hole context assigns the hole type, $\tehole$, to $u$, as in the typing rules. 
However, the first two premises of the expansion subsumption rule, Rule EASubsume, disallow the use of subsumption for holes in analytic position. 
Instead, we have separate analytic rules, {EAEHole} and {EANEHole}, which record the type that the hole is being checked against into the hole context. 
This is again so that we can use type assignment for the internal language --- the type assignment rules TAEHole and TANEHole in Fig.~\ref{fig:hasType} assign a hole closure for $u$ the corresponding type from the hole context.

@@

Empty hole closures, $\dehole{u}{\sigma}{}$,  correspond to the metavariable closures (a.k.a. deferred substitutions) from CMTT, $\cmttclo{u}{\sigma}$.\todo{change notation from CMTT}{} 
We will see how closure environments evolve during evaluation in Sec.~\ref{sec:evaluation}. 
Non-empty hole closures, $\dhole{d}{u}{\sigma}{}$, do not directly correspond to a notion from CMTT (see Sec.~\ref{sec:resumption}).

\subsubsection{Cast Insertion}\label{sec:cast-insertion} 
%
Consider the following example: $\hap{(\halam{x}{\tehole}{\hap{x}{c}})}{c}$. 
The type synthesized for this example viewed as an external expression is $\tehole$, because the hole type annotation on $x$ allows us to apply it as a function of type $\tarr{\tehole}{\tehole}$, as previously discussed, and $c$ can be checked against type $\tehole$ by subsumption. 
However, viewed as an internal expression, this example is not well-typed---we do not have subsumption in the type assignment system defined in Fig.~\ref{fig:hasType}. 
Indeed, it would violate type safety if we could assign a type to this example in the internal language, because beta reduction of this example viewed as an internal expression would result in $c(c)$, which is clearly not well-typed.
The difficulty is that leaving the argument type unknown leaves how the argument is being used (in this case, as a function) also unknown.\footnote{In a system where type reconstruction is first used to try to fill in type holes, we could express a similar example by using $x$ at two or more different types, thereby causing type reconstruction to fail.
%On the other hand, if it is acceptable to arbitrarily choose one of the possible types, and type reconstruction is complete, then type holes will never appear in the internal language and the cast insertion machinery described in this section can be omitted entirely, leaving only the hole closure machinery described previously.
}
\todo{talk about elsewhere? maybe do type-hole-free version of calculus in appendix if there is time}{} 
By our  interpretation of hole types as unknown types from gradual type theory, we can address the problem by performing cast insertion. 
%

The cast form in Hazelnut Live is $\dcasttwo{\dexp}{\htau_1}{\htau_2}$, which, as specified by Rule TACast in Fig.~\ref{fig:hasType}, serves to ``box'' an expression of type $\htau_1$ for treatment as an expression of a consistent type $\htau_2$.%
\footnote{In the earliest work on gradual type theory, the cast form only gave the target type, $\htau_2$ \cite{Siek06a}, but it simplifies matters to include the assigned type, $\htau_1$, in the syntax \cite{DBLP:conf/snapl/SiekVCB15}.}

Casts are inserted during the expansion of function applications and ascriptions. 
The latter is more straightforward: 
Rule~{ESAsc} in Fig.~\ref{fig:expandSyn} inserts a cast from the assigned type to the ascribed  type. 
Theorem~\ref{thm:typed-expansion} inductively ensures that the two types are consistent.  
Note that we included ascription only for the sake of exposition---it can be defined using application together with the half-annotated identity, $e : \tau = \hap{(\halam{x}{\htau}{x})}{e}$, so application expansion, discussed next, is more general.

Cast insertion by the application expansion rule, Rule~{ESAp}, requires a bit more care. 
To understand the rule, consider the expansion for the example above:
\[\hap{\dcasttwo{(\halam{x}{\tehole}{\underline{\hap{\dcasttwo{x}{\tehole}{\tarr{\tehole}{\tehole}}}{\dcasttwo{c}{b}{\tehole}}}})}{\tarr{\tehole}{\tehole}}{\tarr{\tehole}{\tehole}}}{\dcasttwo{c}{b}{\tehole}}
\]
Let us focus on the expansion of the function body, underlined\todo{shaded?}, first. 
A cast has been inserted on both the function expression, $x$, and the the argument, $c$.

The cast on $x$ allows us to treat the variable $x$, which is of type $\tehole$, as being of the matched arrow type $\tarr{\tehole}{\tehole}$, as described in Sec.~\ref{sec:external-statics}. 
The first three premises of Rule~{ESAp} accomplish this by first synthesizing a type for the function expression, here $\tehole$, then determining the matched arrow type, $\tarr{\tehole}{\tehole}$, and then performing analytic expansion on the function expression with this matched arrow type.
The resulting expansion will have some type $\tau_1'$ consistent with the matched arrow type. 
In this case, because $x$ is of variable form, analytic expansion goes through subsumption so $\tau_1'$ is simply $\tehole$. 
The conclusion of the rule inserts the corresponding cast. 
We go through type synthesis then analytic expansion so that the hole context records the matched arrow type for holes in function position, rather than the type $\tehole$ for all such holes as would be the case in a variant of this rule using synthetic expansion for the function expression\todo{put this in the appendix?}{}.

The conclusion of Rule~{ESAp} inserts the cast on the argument's expansion, from the type it is assigned by the final premise of the rule, here $b$ as described at the start of Sec.~\ref{sec:expansion}, to the argument type of the matched arrow type of the function expression, here $\tehole$.

@@

The outer application in the example above goes through the same rule. 
In this case, the cast on the function is the identity cast for $\tarr{\tehole}{\tehole}$. 
We do not attempt to avoid the insertion of identity casts in the core calculus for simplicity (these will simply never fail during evaluation), 
but it is safe to eliminate these during expansion (and some formal accounts of gradual typing do so by defining three application rules, including the original account of \cite{Siek06a}).

@@

The problem is first that \citet{Nanevski2008} defined only the logical reductions for CMTT, viewing it as a proof system for intuitionistic contextual modal logic via the propositions-as-types (Curry-Howard) principle. 
The paper therefore proved only a subject reduction property (which is closely related to type preservation). 
This is not a full dynamic semantics, and in particular, there is no notion of \emph{progress}, i.e. that well-typed terms cannot get ``stuck'' in an undefined state \cite{wright94:_type_soundness}. 
In any case, a conventional dynamic semantics for CMTT would not be immediately relevant to our goal of evaluating incomplete programs because, by our interpretation of hole closures, we would need a dynamic semantics for terms with free metavariables. 
\citet{Nanevski2008} sketched an interpretation of CMTT into the simply-typed lambda calculus with sums under permutation conversion%
\footnote{Permutation conversions are necessary to encode the commuting reductions of CMTT, which in turn are necessary to prove a strong normalization property. These issues are not relevant in \HazelnutLive because, as in the gradually typed lambda calculus, type holes admit non-termination: we can express the Y combinator as $(\halam{x}{\tehole}{x(x)}) (\halam{x}{\tehole}{x(x)})$.}, 
which has been studied by \citet{DBLP:journals/iandc/Groote02}, 
but under this interpretation an analagous problem arises---metavariables become variables of a function type, so again we cannot rely on the standard notion of progress on closed terms.\todo{citations}{}% We also cannot rely on, for example, weak head normalization because \HazelnutLive admits non-termination (due to casts).\todo{citation}

Furthermore, we need to integrate casts into the dynamic semantics. 
Fortunately, the dynamic semantics for the cast calculus from the gradually typed lambda calculus provides most, but not all, of the necessary machinery. 
The first problem is again with progress: in the cast calculus, the only irreducible terms of hole type are casts, which are accounted for by the progress theorem, but in $\HazelnutLive$, holes induce additional irreducible terms. 
The second missing piece is that in prior work on casts, evaluation aborts when cast failure occurs. 
Our goal, as discussed in Sec.~\ref{sec:failed-cast-example}\todo{where?}{}, is for cast failure to instead insert a membrane around the dynamic type error, 
much like a non-empty hole serves as a membrane around a static type error, 
allowing in both cases for evaluation to safely and meaningfully continue past the error when desired.

The dynamic semantics for \HazelnutLive specified in Figures~\ref{fig:isGround}-\ref{fig:step} addresses the difficulties just outlined, 
resulting in a system capable of running incomplete programs without aborting when a hole is encountered. 
This semantics is equipped with a meaningful notion of type safety (involving both preservation and progress). 
We also establish that the standard notion of type safety falls out when running complete terms.

The dynamic semantics is specified as a ``small-step'' transition system. 
The cast-related machinery is based closely on the cast calculus from the ``refined'' account of the gradually typed lambda calculus by \citet{DBLP:conf/snapl/SiekVCB15}, which is known to be theoretically well-behaved. 
In particular, Fig.~\ref{fig:isGround} defines the judgement $\isGround{\htau}$, which distinguishes the base type, $b$, and the least specific arrow type, $\tarr{\tehole}{\tehole}$, as \emph{ground types}, which play a role in simplifying the treatment of function casts. 
Fig.~\ref{fig:isFinal} defines the judgement $\isFinal{d}$, which distinguishes the final forms of the transition system. 
There are two classes of final forms: (possibly-)boxed values and indeterminate forms.%
\footnote{In most accounts of the cast calculus, values and ground types are distinguished with separate grammars together with an implicit identification convention. 
Our judgemental formulation is more faithful to the mechanization and cleaner for our purposes, because we are distinguishing several classes of final forms.}

@@

The judgement $\isBoxedValue{d}$ distinguishes (possibly-)boxed values, which correspond to the values from the cast calculus and include the classic values from the lambda calculus, distinguished by $\isValue{d}$, as well as two cast forms: 
casts between function types (except identity casts) 
and casts from a ground type to the hole type. In both cases, the cast must be on a boxed value.

The judgement $\isIndet{d}$ distinguishes \emph{indeterminate} forms, 
so named because they arise from the presence of expression holes and failed casts in a program, and so, conceptually, their ultimate value awaits programmer action (see Sec.~\ref{sec:resumption}). 
The first two rules specify that hole closures are indeterminate (in the case of non-empty hole closures, when the inner expression is final). 
The rules defining the remaining indeterminate forms are explained simultaneously with the corresponding transition rules below.\todo{relationship to weak head normal forms}
